A fast algorithm for computing best response behaviour strategies in two-person sequential games

### 2014-07-25 paper.tex
implemented primal-dual algo for computing best response strategy

\begin{document}
%\onecolumn
\title{A fast algorithm for computing the best response against the unplanned mixed strategies of an opponent in two-person zero-sum sequential games}

@@

\begin{abstract}
% LESS THAN 200 WORDS !!!!!
% 
% While medical imaging typically provides massive amounts of
% data, the extraction of relevant information in a given
% applicative context remains a difficult challenge.
% XXX : don't call it SPM but just "predictive regions or predictive
% maps"
In this manuscript, we consider the problem of computing the best response against the  unplanned mixed strategies of an opponent
in two-player zero-sum sequential games. The problem considered here encompasses the problem of computing the best response against an opponent's fixed strategy.
The problem considered here encompasses the problem of computing the best response against an opponent's fixed strategy.
An unplanned mixed strategy is simply a strategy chosen from a probability simplex.
The proposed technique employs the primal-dual scheme of A. Chambolle and T. Pock to solve the corresponding saddle-point problem.
The resulting algorithm is simple: all resolvent operators can be effectively computed quasi closed-form, using only elementary algebraic operations.
%%  and is orders of magnitude more efficient than state-of-the-art algorithms like those based on Linear Programming, Interior-Point methods,
%% and more recently, the Nesterov Excessive Gap Technique.
By way of example, we conclude by exhibiting results on Texas Hold'em Poker.
\end{abstract}

\begin{IEEEkeywords}
  game theory; two-person zero-sum sequential game; best response; mixed strategy; simplex; convex-optimization; saddle-point problem; primal-dual scheme
\end{IEEEkeywords}

\section{Introduction: Statement of the problem}
\label{sec:intro}

To begin, let us introduce some technical vocabulary we will be using in this paper. The reader should lookup any standard textbook
on convex optimization for a tutorial introduction to these notions. Viz,\\
$\mathbb{R}_+$: \quad \textit{the nonnegative part of the real-line $\mathbb{R}$}\\
$(x)_+$: \quad \textit{the component-wise maximum of the vector $x$ and 0}\\
$\Delta_{m}$: \quad \textit{the $m-$dimensional probability simplex
%$\{x \in \mathbb{R}_{+}^m\text{ }| \text{ }\textbf{1}^Tx = 1\}$
}\\
$\Pi_C$: \quad \textit{the euclidean projector onto the convex set $C$}\\
$\|K\|$: \quad \textit{the operator norm of a matrix $K$}\\
$F^*$: \quad \textit{the Fenchel transform of a convex l.s.c\footnote{lower semi-contnuous} function $F$}\\
$E^{-1}(U)$: \quad \textit{the pre-image of $U \subset \mathbb{R}^n$ under a matrix $E \in \mathbb{R}^{r,n}$, namely the set $\{x \in \mathbb{R}^n\text{ }| \text{ }Ex \in U\}$}\\\\
Now, consider a two-person zero-sum sequential game with payoff matrix $A \in \mathbb{R}^{m,n}$. We are interested in the problem of
finding the best response strategy $x^* \in \mathbb{R}_{+}^n \cap E^{-1}(\{e\})$ for player 1, given a fixed
(possibly mixed) strategy $y_0$ for player 2, where $E \in \{-1, 0, +1\}^{r,n}$ and $e \in \mathbb{R}^r$ encode player
1's game plan, consistent with the rules of the game. In the language of convex-optimization, one can readily give the following saddle-point formulation for this problem. Viz,

\begin{equation}
  \underset{x \in \mathbb{R}_{+}^n \cap E^{-1}(\{e\})}{min}\text{ }\underset{y \in \Delta_{m}^0}{max}\text{ }y^TAx
\end{equation}

where $\Delta_{m}^0$ is $\Delta_{m}$ if $y_0$ is known, otherwise it is the singleton
%% \footnote{Note that the origin $\textbf{0} \in \mathbb{R}^m$ can be seen a degenerate simplex with radius $0^+$,
%% whilst a singleton  $\{y_0\}$ can be seen as a $y_0-$translation of this simplex.}
$\{y_0\}$.\\\\

In section II, we give a brief overview of existing methods for solving (1). We elaborate our proposed primal-dual algorithm in section III.

\section{Related work}
pending...

\section{The algorithm}
Noting that $i_{E^{-1}(\{e\})}(x) = \underset{\theta \in \mathbb{R}^r}{max}\text{ }\theta^T(e - Ex)$, one can re-write (1) as:
\begin{equation}
  \underset{x \in \mathbb{R}^n}{min}\text{ }\underset{y \in \mathbb{R}^m, \theta \in \mathbb{R}^r}{max}\text{ }
  \left[\begin{array}{@{}cc@{}}y\\\theta\end{array}\right]^TKx + G(x) - F^*(y, \theta)
\end{equation}
where:
\begin{equation}
  \left.
  \begin{aligned}
    K := \left[\begin{array}{@{}cc@{}}A\\-E\end{array}\right]\\
    G:=i_{\mathbb{R}_{+}^n}\\
    F^*: (y, \theta) \mapsto (i_{\Delta_{m}^0}(y) - e^T\theta)
  \end{aligned}
  \right\}
\end{equation}

Furthermore, $F^*$ and $G$ have resolvent (a.k.a \textit{proximity}) operators given by the simple formulae:
\begin{equation}
  \left.
  \begin{aligned}
    (1 + \tau \partial G)^{-1}: x \mapsto \Pi_{\mathbb{R}_{+}^n}(x) = (x)_+\\
    (1 + \sigma \partial F^*)^{-1}: (y, \theta) \mapsto (\Pi_{\Delta_{m}^0}(y), \theta + \sigma e)% (z, \theta) \mapsto \bigl(y, \frac{\theta}{1 + \sigma}\bigr)
  \end{aligned}
  \right\}
\end{equation}
This way, we obtain following primal-dual algorithm for solving (2).

\begin{algorithm}[htb]
\caption{Primal-dual algorithm for computing best response against unplained mixed strategies of opponent}%%  \textbf{Let}
\textbf{Initialize} $\tilde{x}^{(0)} = x^{(0)} \in \mathbb{R}^n$; $y^{(0)} \in \mathbb{R}^{m}$; $\theta^{(0)} \in \mathbb{R}^{r}$;
$\tau, \sigma > 0 \text{ s.t. }\tau\sigma \|K\|^2 \le 1$; $k = 0$.\\
\Repeat{
  convergence}{
  \begin{eqnarray*}
    y^{(k+1)} &\leftarrow& \Pi_{\Delta_{m}^0}\bigl(y^{(k)} + \sigma A \tilde{x}^{(k)}\bigr)\\
    \theta^{(k+1)} &\leftarrow& \theta^{(k)} - \sigma E \tilde{x}^{(k)} + \sigma e\\
    x^{(k+1)} &\leftarrow& \bigl(x^{(k)} - \tau A^Ty^{(k+1)} + E^T\theta^{(k+1)}\bigr)_+\\
    \tilde{x}^{(k+1)} &\leftarrow& 2x^{(k+1)} - x^{(k)}\\
    k &\leftarrow& k + 1
  \end{eqnarray*}
} \Return $x^{(k)}$
\label{Tab:pseudocode_lbfgs}
\end{algorithm}

@@

%% TV$-\ell_1$ regression leads to the following non-smooth convex
%% optimization problem \cite{gramfort-etal:2013a}:
%% \begin{eqnarray}
%%   \mathbf{\hat{w}} := \argmin_{\mathbf{w}}\;\{E(\mathbf{w}) :=
%%   \mathcal{L}(\mathbf{y}, \mathbf{X}, \mathbf{w}) + \alpha
%%   J(\mathbf{w})\},
%%   \label{eq:opt_pb}
%% \end{eqnarray}
%% where $J(\mathbf{w}) := \theta\|\mathbf{w}\|_{\ell_{1}} + (1 -
%% \theta)\|\mathbf{w}\|_{TV}$ is the regularization and $\alpha \ge 0$
%% controls the amount of regularization. The
%% parameter $\theta \in [0, 1]$, also known as the $\ell_{1}$ \textit{ratio}, is the trade-off between the
%% sparsity-inducing penalty $\ell_{1}$ (Lasso) and TV (isotropic Total
%% Variation):
%% % also known as the $\ell_{1}$\emph{-ratio},
%% % XXX = $\ell_{1}$\emph{-ratio} is sklearn jargon
%% \begin{eqnarray}
%% \|\mathbf{w}\|_{\ell_{1}} := \sum_{\omega \in
%%   \Omega}|\mathbf{w}(\omega)|;\;\;\textbf{ }\|\mathbf{w}\|_{TV} :=
%% \sum_{\omega \in \Omega} \|\nabla \mathbf{w}(\omega)\|_{2} \enspace .
%% %% \\
%% %% &=& \sum_{\omega \in
%% %% \Omega}
%% %% \sqrt{\nabla_{x} \mathbf{w}(\omega)^2 + \nabla_{y} \mathbf{w}(\omega)^2 +
%% %%   \nabla_{z}\mathbf{w}(\omega)^2}\\
%% \end{eqnarray}
%% $\mathcal{L}(\mathbf{y}, \mathbf{X}, \mathbf{w})$
%% is the loss function. Here, we focus on the squared loss
%% and the logistic loss, defined \emph{e.g.} in \cite{michel2011}. The
%% squared loss is natural for regression settings, where $\mathbf{y}$ is a
%% continuous variable, but it may also be used for classification
%% \cite{suykens2002}. The logistic loss is harder to optimize, but
%% more natural for classification settings.

@@

%% \paragraph{Uniform approximation by smooth convex surrogates}
%% $\ell_{1}$ (resp. TV) is differentiable almost everywhere,
%% with gradient $\bigl(\textbf{w}(\omega)/|\textbf{w}(\omega|))_{\omega
%%   \in \Omega}$ (resp. $-\dive (\nabla \mathbf{w}/\|\nabla
%% \mathbf{w}\|_2)$), except at voxels $\omega$ of the weights map where
%% \textbf{w}$(\omega) = 0$ (resp. $\|\nabla \mathbf{w}(\omega)\|_{2} =
%% 0$), corresponding to black spots (resp. edges).  A convenient
%% approach (see for example \cite{NESTA, nesterov2005a, nesterov2005b,
%%   beck2012}) for dealing with such singularities is to uniformly
%% approximate the offending function with smooth surrogates that
%% preserve its convexity. Given  % I removed "and other important properties"
%% a smoothing parameter $\mu > 0$, we define \emph{smoothed} versions
%% of $\ell_1$ and TV:
%% %
%% \begin{eqnarray}
%%   \left.
%%   \begin{aligned}
%%     \|\mathbf{w}\|_{\ell_{1},\mu} &:= \sum_{\omega \in \Omega}
%%     \sqrt{\mathbf{w}(\omega)^2 + \mu^2} \quad\\
%%     \|\mathbf{w}\|_{TV, \mu} &:=
%%     \sum_{\omega \in \Omega} \sqrt{\|\nabla \mathbf{w}(\omega)\|_{2}^2 +
%%     \mu^2} \quad
%%     \end{aligned}
%%     \right\}
%% \end{eqnarray}
%% These surrogate upper-bounds are convex and everywhere-differentiable
%% with gradients that are Lipschitz-continuous with constants $1/\mu$ and
%% $\|\nabla\|^2(1/\mu) = 12 / \mu$ respectively.
%% %
%% They lead to smoothed versions of problem \eqref{eq:opt_pb}:
%% \begin{align}
%%   \mathbf{\hat{w}}_{\mu} := \argmin_{\mathbf{w}}\;
%%   \{E_{\mu}(\mathbf{w}) := \mathcal{L}(\mathbf{y}, \mathbf{X}, \mathbf{w}) + \alpha
%%     J_{\mu}(\mathbf{w})\},
%%   \label{eq:sopt_pb}
%% \\
%%     \text{where} \quad
%%     J_{\mu}(\mathbf{w}) :=\theta \|\mathbf{w}\|_{\ell_{1},\mu} +
%%     (1 - \theta)\|\mathbf{w}\|_{TV, \mu} \quad
%% \end{align}

>>>>

\begin{document}
%\onecolumn
\title{A fast algorithm for computing best response behaviour strategies in two-person sequential games}

@@

\begin{abstract}
% LESS THAN 200 WORDS !!!!!
% 
% While medical imaging typically provides massive amounts of
% data, the extraction of relevant information in a given
% applicative context remains a difficult challenge.
% XXX : don't call it SPM but just "predictive regions or predictive
% maps"
In this manuscript, we consider the problem of computing the best response against an opponent's realization plan in two-person sequential games.
The proposed technique employs the primal-dual scheme of A. Chambolle and T. Pock to solve the corresponding contrained convex-optimization problem.
The resulting algorithm is simple: all resolvent operators can be effectively computed in closed-form, using only elementary algebraic operations.
%%  and is orders of magnitude more efficient than state-of-the-art algorithms like those based on Linear Programming, Interior-Point methods,
%% and more recently, the Nesterov Excessive Gap Technique.
By way of example, we conclude by exhibiting results on Texas Hold'em Poker.
\end{abstract}

\begin{IEEEkeywords}
  game theory; two-person sequential game; best response; convex-optimization; contraint; primal-dual scheme
\end{IEEEkeywords}

\section{Introduction}
\label{sec:intro}
\subsection{Notation and Terminology}
To begin, let us introduce some technical notation and terminology we will be using in this paper. The reader should lookup any standard textbook
(for example \cite{sboyd}) on convex optimization for a tutorial introduction to these notions. Viz,\\
\textit{l.s.c}: \quad \textit{acronyme for ``lower semi-continuous''}\\
$\mathbb{R}_+$: \quad \textit{nonnegative part of the real-line $\mathbb{R}$}\\
$i_C$: \quad \textit{indicator function of the convex set $C$}\\
$\Pi_C$: \quad \textit{euclidean projector onto the convex set $C$}\\
$(x)_+$: \quad \textit{component-wise maximum of the vector $x$ and 0}\\
$\|K\|$: \quad \textit{operator norm of the matrix $K$}\\
$F^*$: \quad \textit{Fenchel transform of the convex l.s.c function $F$}\\
$E^{-1}(U)$: \quad \textit{pre-image of $U \subset \mathbb{R}^n$ under a matrix $E \in \mathbb{R}^{p,n}$,
namely the set $\{x \in \mathbb{R}^n\text{ }| \text{ }Ex \in U\}$}

\subsection{Statement of the problem}
Consider a two-person zero-sum sequential game with payoff matrix $A \in \mathbb{R}^{m,n}$. We are interested in the problem of
finding the best response strategy $x^* \in \mathbb{R}_{+}^n \cap E^{-1}(\{e\})$ for player 2, given a fixed
(possibly mixed) strategy $y_0$ for player 1, where $E \in \{-1, 0, +1\}^{p,n}$ and $(1, 0, 0, ..., 0) =: e \in \mathbb{R}^p$.
Recall that $E$ and $e$ encode linear constraints on player 1's ``admissible'' realization plans $x$, in the sequential form representation
of the game (See \cite{bernhard-1996, benhard-agt} for theory on \textit{sequential-form representation}). In the language of convex-optimization,
one can readily give the following saddle-point formulation for this problem. Viz,

\begin{equation}
  \underset{x \in \mathbb{R}_{+}^n \cap E^{-1}(\{e\})}{min}\text{ }-y_0^TAx
\end{equation}

In section II, we give a brief overview of existing methods for solving (1). We elaborate our proposed algorithm in section III.

\section{Related work}
Pending...

\section{The proposed algorithm}
In this section we present the algorithm which is the purpose of this paper, namely an algorithm for solving (1).

\subsection{Derivation of the algorithm}
First observe that (1) can be re-written in the form:
\begin{equation}
  \underset{x \in \mathbb{R}^n}{min}\text{ }{f(-Ax) + g(x)}
\end{equation}
where:\\
\begin{equation}
  \left.
  \begin{aligned}
    g := i_{E^{-1}(\{b\})} + i_{\mathbb{R}_{+}^m}\\
    f: z \mapsto y_0^Tz
    \end{aligned}
  \right\}
\end{equation}

Finally, the primal-dual formulation of this problem is:
\begin{equation}
  \underset{x \in \mathbb{R}^n}{min}\text{ }\underset{y \in \mathbb{R}^m}{max}\text{ }{y^T(-A)x_0 + g(x) - f^*(y))}
\end{equation}

Now, let's observe that $i_{E^{-1}(\{e\})}(x) = \underset{\zeta \in \mathbb{R}^q}{max}\text{ }\zeta^T(e - Ex)$, and so one can re-write (2) as:
\begin{equation}
  \underset{x \in \mathbb{R}^n}{min}\text{ }\underset{y \in \mathbb{R}^m, \zeta \in \mathbb{R}^p}{max}\text{ }
  \begin{bmatrix}y\\\zeta\end{bmatrix}^TKx + G(x) - F^*(y, \zeta)
\end{equation}
where:
\begin{equation}
  \left.
  \begin{aligned}
    K := -\begin{bmatrix}A\\E\end{bmatrix}\\
    G := i_{\mathbb{R}_{+}^n}\\
    F^*: (y, \zeta) \mapsto (i_{\{y_0\}} - \zeta^Te)
  \end{aligned}
  \right\}
\end{equation}

Furthermore, $F^*$ and $G$ have resolvent (a.k.a \textit{proximity}) operators given by the simple formulae:
\begin{equation}
  \left.
  \begin{aligned}
    (1 + \tau \partial G)^{-1} = \Pi_{\mathbb{R}_+^n}: x \mapsto (x)_+\\
    (1 + \sigma \partial F^*)^{-1}: (y, \zeta) \mapsto (y_0, \zeta + \sigma e)
  \end{aligned}
  \right\}
\end{equation}
This way, we obtain following primal-dual algorithm for solving (2).

\begin{algorithm}[htb]
\caption{Primal-dual algorithm for computing best response against opponent's realization plan $y$}%%  \textbf{Let}
\textbf{Initialize} $\tilde{x^{(0)}} = x^{(0)} \in \mathbb{R}^n$; $\zeta^{(0)} \in \mathbb{R}^{q}$;
$\tau, \sigma > 0 \text{ s.t. }\tau\sigma \|K\|^2 < 1$; $k = 0$.\\
\textbf{Precompute} $\eta_0 \leftarrow \tau A^Ty_0$\\
\Repeat{
  convergence}{
  \begin{eqnarray*}
    \zeta^{(k+1)} &\leftarrow& \zeta^{(k)} - \sigma (e - E\tilde{x}^{(k)})\\
    x^{(k+1)} &\leftarrow& \bigl(x^{(k)} + \eta_0 + E^T\zeta^{(k+1)}\bigr)_+\\
    \tilde{x}^{(k+1)} &\leftarrow& 2x^{(k+1)} - x^{(k)}\\
    k &\leftarrow& k + 1
  \end{eqnarray*}
} \Return $x^{(k)}$, $-y_0^TAx^{(k)}$
\label{Tab:pseudocode_lbfgs}
\end{algorithm}

\subsection{Primal-dual gap stopping-criterion and convergence of the algorithm}
Pending...

@@

%% TV$-\ell_1$ regression leads to the following non-smooth convex
%% optimization problem \cite{gramfort-etal:2013a}:
%% \begin{eqnarray}
%%   \mathbf{\hat{w}} := \argmin_{\mathbf{w}}\;\{E(\mathbf{w}) :=
%%   \mathcal{L}(\mathbf{y}, \mathbf{X}, \mathbf{w}) + \alpha
%%   J(\mathbf{w})\},
%%   \label{eq:opt_pb}
%% \end{eqnarray}
%% where $J(\mathbf{w}) := \zeta\|\mathbf{w}\|_{\ell_{1}} + (1 -
%% \zeta)\|\mathbf{w}\|_{TV}$ is the regularization and $\alpha \ge 0$
%% controls the amount of regularization. The
%% parameter $\zeta \in [0, 1]$, also known as the $\ell_{1}$ \textit{ratio}, is the trade-off between the
%% sparsity-inducing penalty $\ell_{1}$ (Lasso) and TV (isotropic Total
%% Variation):
%% % also known as the $\ell_{1}$\emph{-ratio},
%% % XXX = $\ell_{1}$\emph{-ratio} is sklearn jargon
%% \begin{eqnarray}
%% \|\mathbf{w}\|_{\ell_{1}} := \sum_{\omega \in
%%   \Omega}|\mathbf{w}(\omega)|;\;\;\textbf{ }\|\mathbf{w}\|_{TV} :=
%% \sum_{\omega \in \Omega} \|\nabla \mathbf{w}(\omega)\|_{2} \enspace .
%% %% \\
%% %% &=& \sum_{\omega \in
%% %% \Omega}
%% %% \sqrt{\nabla_{x} \mathbf{w}(\omega)^2 + \nabla_{y} \mathbf{w}(\omega)^2 +
%% %%   \nabla_{z}\mathbf{w}(\omega)^2}\\
%% \end{eqnarray}
%% $\mathcal{L}(\mathbf{y}, \mathbf{X}, \mathbf{w})$
%% is the loss function. Here, we focus on the squared loss
%% and the logistic loss, defined \emph{e.g.} in \cite{michel2011}. The
%% squared loss is natural for regression settings, where $\mathbf{y}$ is a
%% continuous variable, but it may also be used for classification
%% \cite{suykens2002}. The logistic loss is harder to optimize, but
%% more natural for classification settings.

@@

%% \paragraph{Uniform approximation by smooth convex surrogates}
%% $\ell_{1}$ (resp. TV) is differentiable almost everywhere,
%% with gradient $\bigl(\textbf{w}(\omega)/|\textbf{w}(\omega|))_{\omega
%%   \in \Omega}$ (resp. $-\dive (\nabla \mathbf{w}/\|\nabla
%% \mathbf{w}\|_2)$), except at voxels $\omega$ of the weights map where
%% \textbf{w}$(\omega) = 0$ (resp. $\|\nabla \mathbf{w}(\omega)\|_{2} =
%% 0$), corresponding to black spots (resp. edges).  A convenient
%% approach (see for example \cite{NESTA, nesterov2005a, nesterov2005b,
%%   beck2012}) for dealing with such singularities is to uniformly
%% approximate the offending function with smooth surrogates that
%% preserve its convexity. Given  % I removed "and other important properties"
%% a smoothing parameter $\mu > 0$, we define \emph{smoothed} versions
%% of $\ell_1$ and TV:
%% %
%% \begin{eqnarray}
%%   \left.
%%   \begin{aligned}
%%     \|\mathbf{w}\|_{\ell_{1},\mu} &:= \sum_{\omega \in \Omega}
%%     \sqrt{\mathbf{w}(\omega)^2 + \mu^2} \quad\\
%%     \|\mathbf{w}\|_{TV, \mu} &:=
%%     \sum_{\omega \in \Omega} \sqrt{\|\nabla \mathbf{w}(\omega)\|_{2}^2 +
%%     \mu^2} \quad
%%     \end{aligned}
%%     \right\}
%% \end{eqnarray}
%% These surrogate upper-bounds are convex and everywhere-differentiable
%% with gradients that are Lipschitz-continuous with constants $1/\mu$ and
%% $\|\nabla\|^2(1/\mu) = 12 / \mu$ respectively.
%% %
%% They lead to smoothed versions of problem \eqref{eq:opt_pb}:
%% \begin{align}
%%   \mathbf{\hat{w}}_{\mu} := \argmin_{\mathbf{w}}\;
%%   \{E_{\mu}(\mathbf{w}) := \mathcal{L}(\mathbf{y}, \mathbf{X}, \mathbf{w}) + \alpha
%%     J_{\mu}(\mathbf{w})\},
%%   \label{eq:sopt_pb}
%% \\
%%     \text{where} \quad
%%     J_{\mu}(\mathbf{w}) :=\zeta \|\mathbf{w}\|_{\ell_{1},\mu} +
%%     (1 - \zeta)\|\mathbf{w}\|_{TV, \mu} \quad
%% \end{align}
